---
title: "STAT 240 Final Group Project Video Presentation"
output: html_document
---

## Discussion Section 318 Group D

## Members 

Katja Zhou

Max Kortebein

Xuancheng Tu

Dewansh Bista

## Introduction

  As the covid-19 cases are continuously rising globally, we are interested in the ratios of different groups of people and their susceptibility to Covid-19, along with the relationship between state population density and state covid-19 death rate (in this report this phrase refers to (death number due to COVID-19) / population)). 

## Thesis statement

  Based on the data we collected, elderly people and males are more likely to die from the coronavirus, while states with bigger population density are more likely to have a higher COVID-19 death rate.
  
## Background

  We have three datasets, covid-19 death data set, population by sex and age data set, population and population density by states dataset. Only the first two will be used for age group and sex analysis.
  
  The Covid-19 dataset, provides the number of deaths due to covid-19 for different age and sex groups. The total population of different sex and age groups are included in the second data set. By these data we can calculate the death rate, which reflects the susceptibility to death caused by covid-19. These three datasets will be joined for data analysis.
  
  
## Graph, Analysis, and Interpretation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
library(tidyverse)
library(lubridate)
source("viridis.R")
source("ggprob.R")
```

```{r,echo=FALSE}
#read in data
covid_raw = read_csv("Provisional_COVID-19_Death_Counts_by_Sex__Age__and_State.csv")
age_sex_raw = read_csv("nc-est2019-alldata-r-file22.csv") 

```

```{r,echo=FALSE}
#data transformation
age_sex_raw <- age_sex_raw %>% 
  filter(MONTH == 10)
age_sex <- age_sex_raw %>% 
  filter(AGE < 999) %>% 
  mutate(age_group = case_when(AGE >= 0 & AGE <= 17 ~ "0-17 years",
                               AGE >= 18 & AGE <= 29 ~ "18-29 years",
                               AGE >= 30 & AGE <= 49 ~ "30-49 years",
                               AGE >= 50 & AGE <= 64 ~ "50-64 years",
                               AGE >= 65 & AGE <= 74 ~ "65-74 years",
                               AGE >= 75 & AGE <= 84 ~ "75-84 years",
                               AGE >= 85 & AGE < 999~ "85 years and over")) %>% 
  group_by(age_group) %>% 
  summarize(total = sum(TOT_POP),
            total_male = sum(TOT_MALE),
            total_female = sum(TOT_FEMALE)) %>% 
  rename(`Age group` = age_group)
age <- age_sex %>% 
  select(`Age group`, total)
sex <- age_sex %>% 
  summarize(Male = sum(total_male),
            Female = sum(total_female))

covid_age <- covid_raw %>% 
  filter(State == "United States") %>% 
  filter(`Age group` == "0-17 years" | `Age group` == "18-29 years" | `Age group` == "30-49 years" | `Age group` == "50-64 years" | `Age group` == "65-74 years" | `Age group` == "75-84 years" | `Age group` == "85 years and over") %>% 
  filter(Sex == "All Sexes") %>% 
  select(`Age group`,`COVID-19 Deaths`) %>% 
  drop_na()  
covid_age = left_join(covid_age,age)

#The focus of this project will be restricted to male and female only
covid_sex <- covid_raw %>%
  filter(State == "United States") %>%
  filter(`Age group` == "All Ages") %>%
  filter(Sex == "Male" | Sex == "Female") %>% 
  select(Sex,`COVID-19 Deaths`) %>%
  drop_na() %>% 
  mutate(total = 0)
for(i in 1:2){
  if(covid_sex[i,"Sex"] == "Male"){
    covid_sex[i,"total"] = pull(sex[1,"Male"])
  }
  if(covid_sex[i,"Sex"] == "Female"){
    covid_sex[i,"total"] = pull(sex[1,"Female"])
  }
}

age_sex2 <- age_sex %>% 
  rename(Male = total_male) %>% 
  rename(Female = total_female) %>% 
  rename(`All Sexes` = total) %>% 
  pivot_longer(c(`All Sexes`,`Male`,`Female`),names_to = "Sex", values_to = "total")
covid_age_sex <- covid_raw %>% 
  filter(State == "United States") %>% 
  filter(Sex != "Unknown") %>% 
  filter(`Age group` == "0-17 years" | `Age group` == "18-29 years" | `Age group` == "30-49 years" | `Age group` == "50-64 years" | `Age group` == "65-74 years" | `Age group` == "75-84 years" | `Age group` == "85 years and over") %>%
  select(`Sex`,`Age group`,`COVID-19 Deaths`) %>% 
  drop_na()
covid_age_sex = left_join(covid_age_sex,age_sex2)
covid_age_sex <- covid_age_sex %>% 
  filter(Sex != "All Sexes")
covid_age2 <- covid_age %>% 
  rename(total_age = `COVID-19 Deaths`) %>% 
  select(-total)
covid_age_sex = left_join(covid_age_sex,covid_age2)
covid_sex2 <- covid_sex %>% 
  rename(total_sex = `COVID-19 Deaths`) %>% 
  select(-total)
covid_age_sex = left_join(covid_age_sex,covid_sex2)

```

## 1. Age group analysis

```{r,echo=FALSE}
total_death = sum(pull(covid_age["COVID-19 Deaths"]))
covid_age_1 <- covid_age %>% 
  mutate(proportion = `COVID-19 Deaths` / total_death)
ggplot(covid_age_1, aes(x = `Age group`, y = proportion, fill = `Age group`)) +
  geom_col() +
  ylab("Proportion of total death") +
  ggtitle("Proportion of total death for each age group") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

This graph represents the proportion of death cases of each age group. The formula of calculation is: (death case of a particular age group) / (total death cases in U.S.). We can see from this graph that elder age groups contribute to higher proportion of covid-19 death cases.

```{r,echo=FALSE}
covid_age_2 <- covid_age %>%
  mutate(death_rate = `COVID-19 Deaths` / total)
## SE for p based on the binomial distribution model
p_binom_se = function(n,p)
{
  return ( sqrt( p*(1-p)/n) )
}
## CI for p using point estimate, se, and normal approximation
p_norm_ci = function(est, se, conf=0.95)
{
  z = qnorm(1 - (1 - conf)/2)
  ci = est + c(-1,1)*z*se
  return( ci )
}
## CI for Agresti-Coull
p_agresti_ci = function(n, x, conf=0.95)
{
  p_tilde = (x+2)/(n+4)
  se = p_binom_se(n+4, p_tilde)
  ci = p_norm_ci(p_tilde, se, conf)
  return( ci )
}
covid_age_3 <- covid_age_2 %>% 
  mutate(p_tilde = (`COVID-19 Deaths` + 2) / (total + 4),
         zscore1 = qnorm(0.025),
         zscore2 = qnorm(0.975),
         se = sqrt(p_tilde*(1-p_tilde)/(total + 4)),
         ci95_1 = p_tilde + zscore1 * se,
         ci95_2 = p_tilde + zscore2 * se)
ggplot(covid_age_3) +
  geom_col(aes(x = `Age group`, y = death_rate, fill = `Age group`)) +
  geom_point(aes(x = `Age group`, y = p_tilde)) +
  geom_segment(aes(x = `Age group`, xend = `Age group`, y = ci95_1, yend = ci95_2), size = 1.0) +
  ylab("covid-19 death rate") +
  ggtitle("COVID-19 death rate (COVID-19 death number) / (population)\nfor different age group with Agresti-Coull 95% percent confidence \ninterval") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

This graph represents the death rate of each age group.  The calculation formula is (COVID-19 death number of a particular age group) / (total population of that age group in U.S.). Confidence intervals are small so it is hard to visualize, so exact value in decimal numbers are shown in the table below. In the table, p_tilde means the estimated death rate of the corresponding age group calculated using Agresti-Coull method. We can see from the graph and data table that elder age groups have higher covid-19 death rate.


```{r,echo=FALSE}
covid_age_3 <- covid_age_3 %>% 
  select(`Age group`, "death_rate", ci95_1, ci95_2) %>% 
  mutate(death_rate = format(death_rate, scientific = FALSE),
         ci95_1 = format(ci95_1, scientific = FALSE),
         ci95_2 = format(ci95_2, scientific = FALSE),) %>% 
  rename("95%_conf_intvl_1" =ci95_1,
         "95%_conf_intvl_2" =ci95_2,
         "p_tilde" = death_rate) 
covid_age_3
```



Likelihood ratio test

1. State hypothesis

$H_0$: $p_1$ = ... = $p_7$ i.e., covid-19 death rates of different age groups are actually the same. 

$H_1$: not $p_1$ = ... = $p_7$ i.e., covid-19 death rates of different age groups are not the same. 

2.Test statistic

$lrt$ = $2(ln L_1-lnL_0)$

3.Calculate p-value

```{r,echo=FALSE}
covid_age_4 <- covid_age %>% 
  mutate(p_0 = sum(`COVID-19 Deaths`)/sum(total)) %>%
  mutate(phat = `COVID-19 Deaths` / total) %>% 
  mutate(log_L0 = dbinom(`COVID-19 Deaths`,total,p_0,log=TRUE),
         log_L1 = dbinom(`COVID-19 Deaths`,total,phat,log=TRUE))
lrt = covid_age_4 %>% 
  summarize(ln_L0 = sum(log_L0),
            ln_L1 = sum(log_L1),
            lrt = 2*(ln_L1 - ln_L0),
            R = exp(ln_L0-ln_L1))
lrt = lrt %>%
  mutate(p_value = 1 - pchisq(lrt,6)) %>% 
  select(-R)
lrt
```

p-value is less than 0.05 (not exactly 0, but too close to 0 which is unable to be precisely presented by computer), which is statistically significant.

## 2. Sex analysis


```{r,echo=FALSE}
total_death = sum(pull(covid_sex["COVID-19 Deaths"]))
covid_sex_1 <- covid_sex %>% 
  mutate(proportion = `COVID-19 Deaths` / total_death)
ggplot(covid_sex_1, aes(x = Sex, y = proportion, fill = Sex)) +
  geom_col() +
  ylab("Proportion of total death") +
  ggtitle("Proportion of total death for each sex group") 
```
This graph represents the proportion of death cases of male and female. The formula of calculation is: (death case of a particular sex group) / (total death cases in U.S.). We can see from the graph that males contribute to larger proportion of covid-19 death cases than females do.


```{r,echo=FALSE}
covid_sex_2 <- covid_sex %>%
  mutate(death_rate = `COVID-19 Deaths` / total)
covid_sex_3 <- covid_sex_2 %>% 
  mutate(p_tilde = (`COVID-19 Deaths` + 2) / (total + 4),
         zscore1 = qnorm(0.025),
         zscore2 = qnorm(0.975),
         se = sqrt(p_tilde*(1-p_tilde)/(total + 4)),
         ci95_1 = p_tilde + zscore1 * se,
         ci95_2 = p_tilde + zscore2 * se)
ggplot(covid_sex_3) +
  geom_col(aes(x = Sex, y = death_rate, fill = Sex)) +
  geom_point(aes(x = Sex, y = p_tilde)) +
  geom_segment(aes(x = Sex, xend = Sex, y = ci95_1, yend = ci95_2), size = 1.0) +
  ylab("covid-19 death rate") +
  ggtitle("COVID-19 death rate for different sex group")
```


This graph represents the death rate of each sex group.  The calculation formula is (COVID-19 death number of a particular sex group) / (total population of that sex group in U.S.). Confidence intervals are small so it is hard to visualize. Exact values of confidence intervals are shown in the table below. In the table, p_tilde means the estimated death rate of the corresponding age group calculated using Agresti-Coull method. We can see from the graph and data table that compared to females, males have higher coid-19 death rate. 


```{r,echo=FALSE}
covid_sex_3 <- covid_sex_3 %>% 
  select(`Sex`, "death_rate", ci95_1, ci95_2) %>% 
  mutate(death_rate = format(death_rate, scientific = FALSE),
         ci95_1 = format(ci95_1, scientific = FALSE),
         ci95_2 = format(ci95_2, scientific = FALSE),) %>% 
  rename("95%_conf_intvl_1" =ci95_1,
         "95%_conf_intvl_2" =ci95_2,
         "p_tilde" = death_rate) 
covid_sex_3
```

Likelihood ratio test

1. State hypothesis

$H_0$: $p_1$ = $p_2$ i.e., the covid-19 death rates of male and female are the same.

$H_1$: $p_1$ ≠ $p_2$ i.e., the covid-19 death rated of male and female are not the same.

2.Test statistic

$lrt$ = $2(ln L_1-lnL_0)$

3.Calculate p-value

```{r,echo=FALSE}
covid_sex_4 <- covid_sex %>% 
  mutate(p_0 = sum(`COVID-19 Deaths`)/sum(total)) %>%
  mutate(phat = `COVID-19 Deaths` / total) %>% 
  mutate(log_L0 = dbinom(`COVID-19 Deaths`,total,p_0,log=TRUE),
         log_L1 = dbinom(`COVID-19 Deaths`,total,phat,log=TRUE))
lrt = covid_sex_4 %>% 
  summarize(ln_L0 = sum(log_L0),
            ln_L1 = sum(log_L1),
            lrt = 2*(ln_L1 - ln_L0),
            R = exp(log_L0-log_L1))
lrt = lrt %>%
  mutate(p_value = 1 - pchisq(lrt,1)) %>% 
  select(-R)
lrt[-c(2),]
```

p-value is less than 0.05 (not exactly 0, but too close to 0 which is unable to be precisely presented by computer), which is statistically significant.

## 3. Analysis combining sex and age group

```{r,echo=FALSE}
covid_age_sex1 <- covid_age_sex %>% 
  mutate(sexdeath_proportion = `COVID-19 Deaths` / total_sex) %>% 
  mutate(agedeath_proportion = `COVID-19 Deaths` / total_age) %>% 
  mutate(death_rate = `COVID-19 Deaths` / total)
```

```{r,echo=FALSE}
ggplot(covid_age_sex1, aes(x = `Age group`, y = sexdeath_proportion, fill = `Age group`)) +
  geom_col() +
  ylab("Proportion of total death\nwithin different sex groups") +
  facet_wrap(~Sex) +
  ggtitle("Proportion of total death for each age group\nwithin different sex groups") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

This graph represents proportion of total death for each age group within different sex groups. The calculation formula is (COVID-19 death number of a particular age group with a certain sex) / (total COVID-19 death number of that sex). We can see from this graph that in most cases, elder age groups contribute to higher proportion of covid-19 death cases, both for male and female. However, there is a counter example, which is 85 years and over male group. It is the eldest age group, but it is not the age group with highest proportion of covid-19 death cases among males. Instead, 75-84 years old age group has the highest proportion of covid-19 death cases for males. 


```{r,echo=FALSE}
ggplot(covid_age_sex1, aes(x = Sex, y = agedeath_proportion, fill = Sex)) +
  geom_col() +
  ylab("Proportion of total death\nwithin different age groups") +
  facet_wrap(~`Age group`) +
  ggtitle("Proportion of total death for each sex group\nwithin different age groups") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

This graph represents proportion of total death for each sex group within different age groups. The calculation formula is (COVID-19 death number of a particular sex group within a certain age group) / (total COVID-19 death number of that age group). We can see from the graph that for most age groups, males contribute to larger proportion of covid-19 death cases than females do. However, for age group of 85 years and over, there are more female death cases than male death cases.

```{r,echo=FALSE}
age_sex3 <- age_sex %>% 
  mutate(`%_of_total_male` = total_male / sum(total_male),
         `%_of_total_female` = total_female / sum(total_female),
         `%_of_total` = total / sum(total))
age_sex3 %>% 
  print(width = Inf)
```

This is the table of United States population distribution. We can see that for the age group of 85 years and over, number of female is significantly greater than male. Also, males of 85 years and over have a smaller porportion of total male population (1.52%) than females of 85 years and over who have a proportion of 2.58% of total female population, while the overall proportion of people of 85 years over is 2.06%. This might explain the exception above, and will be further discussed in discussion section.

```{r,echo=FALSE}
ggplot(covid_age_sex1, aes(x = `Age group`, y = death_rate, fill = `Age group`)) +
  geom_col() +
  ylab("COVID-19 death rate") +
  facet_wrap(~`Sex`) +
  ggtitle("COVID-19 death rate of different age groups with different sexes") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

This graph represents the COVID-19 death rate of different age groups with different sexes. This graph is faceted by sex. The calculation formula is (COVID-19 death number of a particular age group with a certain sex) / (total population of that age group with that sex). We can see from the graph that for both male and female, elder age groups have higher covid-19 death rate, which is consistent with what we found previously in age group analysis. 

For the next two graphs, the first graph will represent data of people under 49 years old; the other graph will represent people above 49 years old. Creating these two separated graphs instead of one graph is for better data visualizaion , with no statistical meaning.

```{r,echo=FALSE}
covid_age_sex2 <- covid_age_sex1 %>% 
  filter(`Age group` == "0-17 years" | `Age group` == "18-29 years" | `Age group` == "30-49 years")
ggplot(covid_age_sex2, aes(x = `Sex`, y = death_rate, fill = `Sex`)) +
  geom_col() +
  ylab("COVID-19 death rate") +
  facet_wrap(~`Age group`) +
  ggtitle("COVID-19 death rate of different sex groups\n within different age groups (under 49 years old)") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```
```{r,echo=FALSE}
covid_age_sex3 <- covid_age_sex1 %>% 
  filter(`Age group` == "50-64 years" | `Age group` == "65-74 years" | `Age group` == "75-84 years" | `Age group` == "85 years and over")
ggplot(covid_age_sex3, aes(x = `Sex`, y = death_rate, fill = `Sex`)) +
  geom_col() +
  ylab("COVID-19 death rate") +
  facet_wrap(~`Age group`) +
  ggtitle("COVID-19 death rate of different sex groups\n within different age groups (above 49 years old)") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

These two graphs represent the COVID-19 death rate of different sex groups within different age ranges. This graph is faceted by age groups. The calculation formula is (COVID-19 death number of a particular sex group within a certain age group) / (total population of that sex group within that age group). We can see from these two graphs that for all age groups, compared to females, males have higher coid-19 death rate, which is consistent with what we found previously in sex analysis.

  
## Conclusion

  We found that people of elder age groups are more vulnerable to covid-19 since the covid-19 death rate is higher for age groups that are older. We also found that males are more vulnerable to covid-19 than females, since the covid-19 death rate is higher for males, compared to females. 
  
## Reference

^[Provisional_COVID-19_Death_Counts_by_Sex__Age__and_State.csv. C.D.C. Retrieved from https://data.cdc.gov/NCHS/Provisional-COVID-19-Death-Counts-by-Sex-Age-and-S/9bhg-hcku Retrieved date: November 4th, 2020]

^[nc-est2019-alldata-r-file22.csv. United States Census Bureau. Retrieved from https://www.census.gov/data/tables/time-series/demo/popest/2010s-national-detail.html#par_textimage_98372960]

^[csvData.csv. World Population Review. Retrieved from https://worldpopulationreview.com/state-rankings/state-densities]